---
title: "PaTrEco - EXERCISE 2"
author: "Katjusa Mezek"
date: "5/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Patterns and trends in environmental data - Exercise 2

## Packages & data import

```{r echo=T, results='hide',warning = FALSE, message = FALSE}
# Load packages

library(readr)        # to import tabular data (e.g. csv)
library(dplyr)        # to manipulate (tabular) data
library(ggplot2)      # to visualize data
library(sf)           # to handle spatial vector data
library(terra)        # To handle raster data
library(lubridate)
library(zoo)
library(scales)

# Import the Wildschwein data

wildschwein_BE <- read_delim("wildschwein_BE_2056.csv",",") # adjust path

wildschwein_BE <- st_as_sf(wildschwein_BE, coords = c("E", "N"), crs = 2056, remove = FALSE)

```

# Task 1 - Getting an overview

```{r echo=T, results='hide',warning = FALSE, message = FALSE}
# Calculate the time difference between subsequent rows

wildschwein_BE <- group_by(wildschwein_BE,TierID) # group by TierID

wildschwein_BE$timelag  <- as.numeric(difftime(lead(wildschwein_BE$DatetimeUTC), wildschwein_BE$DatetimeUTC),units = "secs")

# turn timelag to integer

wildschwein_BE$timelag = as.integer(wildschwein_BE$timelag)

# A summary for individuals
wildschwein_BE %>% 
  group_by(TierID) %>%
  summarise(
    mean_timelag=mean(timelag, na.rm=T)
  )

```

## Questions
1.) How many individuals were tracked?
```{r}
unique(wildschwein_BE$TierName)
unique(wildschwein_BE$TierID)

#three individuals were tracked
```

2.) For how long were the individual tracked? Are there gaps?
```{r}
ggplot(wildschwein_BE, aes(x=wildschwein_BE$DatetimeUTC, y=wildschwein_BE$TierID)) +
  geom_line()

range_002A=range(wildschwein_BE$DatetimeUTC[wildschwein_BE$TierID=="002A"])
range_016A=range(wildschwein_BE$DatetimeUTC[wildschwein_BE$TierID=="016A"])
range_018A=range(wildschwein_BE$DatetimeUTC[wildschwein_BE$TierID=="018A"])

# The animal 002A was tracked longer (Oct2014 - Jul2015) than 016A and 018A.
# The start of the tracking is different for each animal (same day - 2014-11-07, but different time).
#The tracking ended on different dates, as seen in the plot. The tracking of 018A and 022A ended on the same day but not same time

```

3.) Were all individuals tracked concurrently or sequentially?
```{r}
# The individuals were tracked concurrently because there is a large time area when they were all tracked at the same time between January 2015 and June 2015), but with slight delays.

```

4.) What is the temporal sampling interval between the locations?
```{r}
# The sampling interval can be interpreted as the time lag
# it is approximately 900s (15 minutes)
```

# Task 2 - Deriving movement parameters I: Speed
```{r echo=T, results='hide',warning = FALSE, message = FALSE}
# Calculate Euclidean distance
euc.dist = function(x,y) (sqrt((lead(x)-x)^2+(lead(y)-y)^2))

wildschwein_BE$steplength= euc.dist(wildschwein_BE$E, wildschwein_BE$N)

# Calculate speed between locations
wildschwein_BE$Speed = wildschwein_BE$steplength/wildschwein_BE$timelag

#The unit is in degrees/s (timelag is in seconds and steplength is in degrees, since it was calculated from the euclidean distance)
#In order to get measurable distance we would need to take into account the earths curve and terrain - haversine

```

# Task 3 - Cross-scale movement analysis

```{r echo=T, echo=T, results='hide',warning = FALSE, message = FALSE}

# data import
caro <- read_delim("caro60.csv",",")
caro <- st_as_sf(caro, coords = c("E", "N"), crs = 2056, remove = FALSE)

#reduce granularity of sampling interval by selecting every 3, 6 and 9th row 
caro_3 = caro[seq(from = 1, to = 200, by = 3),] 
caro_6 = caro[seq(from = 1, to = 200, by = 6),]
caro_9 = caro[seq(from = 1, to = 200, by = 9),]

nrow(caro)
nrow(caro_3)
nrow(caro_6)
nrow(caro_9)

#calculate timelag, steplength and speed for all caro datasets

#caro
caro$timelag  <- as.numeric(difftime(lead(caro$DatetimeUTC), caro$DatetimeUTC),units = "secs")
caro$steplength= euc.dist(caro$E, caro$N)
caro$Speed = caro$steplength/caro$timelag

#caro_3
caro_3$timelag  <- as.numeric(difftime(lead(caro_3$DatetimeUTC), caro_3$DatetimeUTC),units = "secs")
caro_3$steplength= euc.dist(caro_3$E, caro_3$N)
caro_3$Speed = caro_3$steplength/caro_3$timelag

# caro_6
caro_6$timelag  <- as.numeric(difftime(lead(caro_6$DatetimeUTC), caro_6$DatetimeUTC),units = "secs")
caro_6$steplength= euc.dist(caro_6$E, caro_6$N)
caro_6$Speed = caro_6$steplength/caro_6$timelag

# caro_9
caro_9$timelag  <- as.numeric(difftime(lead(caro_9$DatetimeUTC), caro_9$DatetimeUTC),units = "secs")
caro_9$steplength= euc.dist(caro_9$E, caro_9$N)
caro_9$Speed = caro_9$steplength/caro_9$timelag

#make one data frame
caro_13=rbind(caro,caro_3)
caro_13$Trajectory=c(replicate(200, "1 minute"), replicate(67,"3 minutes"))

caro_16=rbind(caro,caro_6)
caro_16$Trajectory=c(replicate(200, "1 minute"), replicate(34,"6 minutes"))

caro_19=rbind(caro,caro_9)
caro_19$Trajectory=c(replicate(200, "1 minute"), replicate(23,"9 minutes"))

caro_all=rbind(caro, caro_3 , caro_6, caro_9)
caro_all$Trajectory=c(replicate(200, "1 minute"), replicate(67,"3 minutes"), replicate(34,"6 minutes"),replicate(23,"9 minutes")  )

# visual comparisons to original
#plot caro 3 compared to caro
ggplot(caro_13, aes(y=N, x=E, col=Trajectory) ) +
  geom_path()+ geom_point()

#plot caro 6 compared to caro
ggplot(caro_16, aes(y=N, x=E, col=Trajectory) ) +
  geom_path()+ geom_point()

#plot caro 9 compared to caro
ggplot(caro_19, aes(y=N, x=E, col=Trajectory) ) +
  geom_path()+ geom_point()

#plot all caro datasets with different granularities
ggplot(caro_all, aes(y=N, x=E, col=Trajectory) ) +
  geom_path()+ geom_point()

#compare derived speed at different sampling intervals
#preprocessing 
caro_all$Time <- format(as.POSIXlt(caro_all$DatetimeUTC, "%Y-%m-%d %H:%M:%S", tz = "UTC"), format = "%H:%M")
str(caro)

ggplot(caro_all, aes(Time,Speed, col = Trajectory, group = 1)) + geom_line()

#fix y scale
caro_all$Time <- format(as.POSIXct(caro_all$DatetimeUTC, "%Y-%m-%d %H:%M:%S", tz = "UTC"), format = "%H:%M")
caro_all$Time = strptime(caro_all$Time, format = "%H:%M" )
caro_all$Time = as.POSIXct(caro_all$Time, format = "%H:%M")

str(caro_all)

# plot derived speed at different sampling intervals
ggplot(caro_all, aes(Time,Speed, col = Trajectory, group = 1)) + geom_line() +
  scale_x_datetime(breaks = date_breaks("1 hour"), labels=date_format("%H:%M:%S"))+
  theme_minimal() + ggtitle("Comparing derived speed at different sampling intervals")
 
```

The plots in this exercise indicate that granularity effects the results. The lineplot shows the speeds at different granularities (1,3,6 and 9 minutes). The higher granularity line (eg. sampling every 1 minute) shows ore noisiness while lower granularity (eg. every 9 minutes) shows allot less noisiness in the data. If the granularity is too low (9 minutes) some peaks are lost, as seen in the plot. The same is seen in the plotted trajectory maps. 

# Task 4 - Deriving movement parameters II: Rolling window functions
```{r,echo=T, results='hide',warning = FALSE, message = FALSE}
# demo 
example <- rnorm(10)
rollmean(example,k = 3,fill = NA,align = "left")
##  [1]  0.93634335  0.31709038  0.02370048  0.67869801  0.73369105  0.50401344
##  [7] -0.56144365 -0.56902598          NA          NA
rollmean(example,k = 4,fill = NA,align = "left")
##  [1]  0.6775521  0.2045005  0.5848215  0.5255629  0.3446928  0.1459635
##  [7] -0.4102301         NA         NA         NA

# A higher k - smoother data

# Run rollmean on caro
caro$Speedk3=rollmean(caro$Speed,k = 3,fill = NA,align = "left")
caro$Speedk5=rollmean(caro$Speed,k = 5,fill = NA,align = "left")

# Make long data
library(tidyr)
caro_long <- gather(caro, key = k_value , value = Speed, Speed:Speedk5, factor_key = TRUE) 

caro_long$Time <- format(as.POSIXlt(caro_long$DatetimeUTC, "%Y-%m-%d %H:%M:%S", tz = "UTC"), format = "%H:%M")

# Data preprocessing
caro_long$k_value=as.factor(caro_long$k_value)
caro_long$Time = strptime(caro_long$Time, format = "%H:%M" )
caro_long$Time = as.POSIXct(caro_long$Time, format = "%H:%M")

#visualize speed
ggplot(caro_long, aes(Time, Speed,col=k_value, group = 1)) + geom_line()+
  scale_x_datetime(breaks = date_breaks("1 hour"), labels=date_format("%H:%M:%S"))+
  theme_minimal() + ggtitle("Comparing different window sizes (k-values)")
```
